---
title: "Dimensionality Reduction using Spectral Analysis"
author: Sebasti\'{a}n Mart\'{i}nez
output: md_document
---
# Dimensionality Reduction using Spectral Analysis

This short document presents an interesting alternative to Principal Component Analysis by using spectral decomposition. The original content of this document is based on a [presentation](https://juanitorduz.github.io/laplacian_eigenmaps_dim_red.html) given by Dr. Juan Orduz at the 2018 PyData Berlin conference. The linked post is a practical application using Python to perform non-linear dimensionality reduction based on spectral techniques. This post aims at replicating the same results, but using R instead. 

This is a list of all the packages you are going to need to carry out this exercise in R. 

* <code>loe</code>: Carries out the main calculations of the spectral analysis. Functions used: So and so
* <code>igraph</code>: Graph analysis. Functions used: so and so
* <code>ggplot2</code>: Plotting package. Carries out the main calculations of the spectral analysis
* <code>scatterploted</code>: 3D plotting package. 
* <code>reshape</code>: Tydiverse package used to change the shape of dataframes. Functions used: so and s




```{r setup, include=FALSE}
#knitr::opts_chunk$set(echo = TRUE)
if(!require(dimRed)) install.packages("dimRed",repos = "http://cran.us.r-project.org")
library("dimRed")
if(!require(diffusionMap)) install.packages("diffusionMap",repos = "http://cran.us.r-project.org")
library("diffusionMap")

if(!require(Graph)) install.packages("igraph",repos = "http://cran.us.r-project.org")
library("igraph")

if(!require(loe)) install.packages("loe",repos = "http://cran.us.r-project.org")
library("loe")

if(!require(ggplot2)) install.packages("ggplot2",repos = "http://cran.us.r-project.org")
library("ggplot2")

if(!require(scatterplot3d)) install.packages("scatterplot3d",repos = "http://cran.us.r-project.org")
library("scatterplot3d")


if(!require(FNN)) install.packages("FNN",repos = "http://cran.us.r-project.org")
library("FNN")

if(!require(reshape)) install.packages("reshape",repos = "http://cran.us.r-project.org")
library("reshape")

if(!require(micEcon)) install.packages("micEcon",repos = "http://cran.us.r-project.org")
library("micEcon")
```

*NOTE:* As this is only a "translation" post, we are going to follow the same examples from the original. 

## The Shortest Algebra Section Ever
Consider $M_{nxn}(\mathbb{R})$ the space of all $nxn$ matrices with real elements. Let $A\inM_{nxn}(\mathbb{R})$ be a symmetric matrix.
$\lambda \in \mathbb{C}$ is said to be an *eigenvalue* of A, with associated *eigenvector* $f\in{\mathbb{R}}^n,\ f\neq0$, if $$Af = \lambda f$$
Additionally, we say that a set of vectors $\mathcal{B} = \{f_1, f_2, \cdots , f_n\}$ is a basis for $\mathbb{R}^n$ if 
1. They are linearly independent; and 2.  They generate $\mathbb{R}^n$
And we can say that $\mathcal{B}$ is an orthonormal basis if for two elements $f_i, f_j \in \mathcal{B}$, $<f_i,fj} = \delta_{ij}$.

The Spectral Theorem, at the center of this entire post, suggests that there exists an orthonormal basis for $\mathbb{R}^n$ consisting of the eigenvalues of $A$, where each eigenvalue is real. 

## Toy Example
Consider the adjacency matrix from the toy example:

```{r}
adj_mat <- matrix(c(0,1,1,1,1,0,0,0,1,0,0,0,1,0,0,0), nrow = 4, ncol = 4)
adj_mat
``` 

We can create the graph associated to this matrix by using the <code>graph_from_adjacency_matrix</code> function from the <code>igraph</code> package. We also make the degree matrix, which places the degree of each one of the nodes in the graph in its diagonal, and the Laplacian matrix which calculates the difference between the degree matrix and the adjacency matrix. 

```{r, echo = TRUE}
# Setting the graph from the adjacency matrix
g <- igraph::graph_from_adjacency_matrix(adjmatrix = adj_mat)

# Giving it a name
g <- set_graph_attr(g, "name", "Toy Example")

# Plotting the graph with smaller arrow sizes, and a title
igraph::plot.igraph(x = g, edge.arrow.size = 0.5, main = "Toy Example")


# Degree matrix and Laplacian Matrix
deg_mat <- Diagonal(rowsum(adj_mat,group = c(1,1,1,1)), n = 4)
laplacian <- as.matrix(deg_mat-adj_mat)
```



We are now going to check what this projection of this two dimensional object onto $\mathbb{R}$ looks like using the <code>spec.emb</code> function from the <code>loe</code> package. We need to include the number of dimensions of the desired space in which we want to project the graph. In this case, $y$ contains the 

```{r, echo = TRUE}
y <- loe::spec.emb(adj_mat,1)
y
```

Using what we learned from the slides and the presentation available in the original post, we would like to check if, analitically, the result given by <code>loe::spec.emb</code> is what it should be.

Recall that the first eigenvalue for our matrix was 1. 
Using $y$






```{r}
lambda_1 <- 1

a1 <- laplacian %*% y
a2 <- lambda_1*(deg_mat %*% y)
result <- a1 - a2

```


```{r}

#make sphere in R

number_points <- 1000
r <- runif(number_points, min = 0, max = 2*pi)
rho <- runif(number_points, min = 0, max = 2*pi)

x <- sin(rho)*cos(r)
y <- sin(rho)*sin(r)
z <- cos(rho)
data <- as.data.frame(matrix(c(x, y, z), ncol =3))
colnames(data) <- c("x_ax","y_ax", "z_ax")


x <- data$x_ax[!(abs(data$z_ax)>0.99)]
y <- data$y_ax[!(abs(data$z_ax)>0.99)]
z <- data$z_ax[!(abs(data$z_ax)>0.99)]

data_nopo <- as.data.frame(matrix(c(x, y, z), ncol =3))
colnames(data_nopo) <- c("x_ax","y_ax", "z_ax")
x1 <- rnorm(length(data_nopo$x_ax))

s_colors <- rgb(abs(data_nopo$x_ax), abs(data_nopo$y_ax), abs(data_nopo$z_ax), maxColorValue = 1)
scatterplot3d(data_nopo, pch = 16, main="Sphere",
              xlab = "X",
              ylab = "Y",
              zlab = "Z",
              box = FALSE,
              color = s_colors)

```

```{r}

ggplot2::ggplot(data = data_nopo, aes(x = x_ax, y = y_ax)) + geom_point( color = s_colors)

n_components <- 2
prin <- prcomp(x = data_nopo, scale. = T, rank. = n_components)
#prin$center
#biplot(prin)
prinx <- prin$x[,1]
priny <- prin$x[,2]
prin_comp <- as.data.frame(matrix(c(prinx, priny), ncol = 2))
colnames(prin_comp) <- c("first", "second")

ggplot2::ggplot(data = prin_comp, aes(x = first, y = second)) + geom_point( color = s_colors)

n_neighbors <- 50

# Fit the object.
nn <- get.knn(data = data_nopo, k = n_neighbors)$nn.index
nn <- as.matrix(nn)
nn <- as.data.frame(nn)
nn$id <- rownames(nn)


nn_melt <- melt(nn, id.vars = c("id"))
nn_melt <- melt(nn)
nn_melt <- nn_melt[c("id", "value")]
nn_melt$id <- as.integer(nn_melt$id)


g <- graph.edgelist(as.matrix(nn_melt), directed=TRUE)
#plot(g)

g <- graph.edgelist(as.matrix(nn_melt), directed=FALSE)
#plot(g)

adj_mat <- as.data.frame(as.matrix(get.adjacency(graph = g)))
adjmat <- ifelse(adj_mat == 2, 1, ifelse(adj_mat == 1, 1, 0))


spec_emb <- spec.emb(A = as.matrix(adjmat), p = n_components)
spec1 <- spec_emb[,1]
spec2 <- spec_emb[,2]
spec <- as.data.frame(matrix(c(spec1, spec2), ncol = 2))
colnames(spec) <- c("first", "second")


ggplot2::ggplot(data = spec, aes(x = first, y = second)) + geom_point( color = s_colors)

```

